Highly Available Transactions: Virtues and Limitations\\(Extended Version)

### 2013-07-31 writeup/hat-appendix.tex
Last round of comments from Alan and Aaron

To reason about isolation anomalies, we use Adya's concept of a
conflict graph, which is composed of dependencies between
transaction. The definitions in this section are directly from Adya,
with two difference. First, we expand Adya's formalism to deal
with \textit{per-item} dependencies. Second, we define session
dependencies (Definitions~\ref{def:sd1} and~\ref{def:sd2}).

@@

\begin{definition}[Directly Read-Depends]
$T_j$ directly read-depends on transaction $T_i$ if it directly
item-read-depends or directly predicate-read-depends on $T_i$.
\end{definition}

@@

\begin{definition}[Directly Anti-Depends]
Transaction $T_j$ directly anti-depends on transaction $T_i$ if it
directly item-anti-depends or directly predicate-anti-depends on
$T_i$.
\end{definition}

@@

\begin{definition}[Directly Write-Depends]
A transaction $T_j$ directly write-depends on transaction $T_i$ if
$T_i$ directly write-depends by $x$ on $T_j$ for some item $x$.
\end{definition}

@@

The dependencies for a history $H$ form a graph called its Directed
Serialization Graph ($DSG(H)$). If $T_j$ directly write-depends on
$T_i$ by $x$, we draw $T_i\overset{ww_x}\longrightarrow T_j$. If $T_j$
read-depends on $T_i$ by $x$, we draw
$T_i \overset{rw_x}\longrightarrow T_j$. If $T_j$ directly
anti-depends on transaction $T_j$ by $x$, we draw
$T_i \overset{wr_x}\DashedArrow T_j$. If $T_j$ session-depends on
$T_i$ in session $S$, we draw $T_i \overset{S}\longrightarrow T_j$.

@@

\begin{definition}[Item-Many-Preceders (N-I-CI)]
\label{def:n-ici}
A history $H$ exhibits phenomenon N-ICI if $DSG(H)$ contains a
transaction $T_i$ such that $T_i$ directly item-read-depends by $x$ on more
than one other transaction.
\end{definition}

\begin{figure}[H]
\begin{align*}
\small
T_1 &: w_x(1)\\
T_2 &: w_x(2)\\
T_3 &: r_x(1)~r_x(2)
\end{align*}
\caption{Example of \textit{N-ICI} anomaly.}
\label{fig:nici-history}
\end{figure}

@@

\begin{definition}[Item Cut Isolation (I-CI)]
A system that provides Item Cut Isolation prohibits phenomenon N-I-CI.
\end{definition}

\begin{definition}[Predicate-Many-Preceders (N-P-CI)]
A history $H$ exhibits phenomenon N-I-CI if, for all predicate-based
reads $r_i(P_i:Vset(P_i))$ and $r_j(P_j:Vset(P_j)$ in $T_k$ such that
the logical ranges of $P_i$ and $P_j$ overlap (call it $P_o$), the set
of transactions that change the matches of $P_o$ for $r_i$ and $r_j$
differ.
\end{definition}

\begin{definition}[Predicate Cut Isolation (P-CI)]
A system that provides Predicate Cut Isolation prohibits phenomenon N-P-CI.
\end{definition}

\begin{definition}[Observed Transaction Vanishes (N-RA)]
A history $H$ exhibits phenomenon N-RA if $USG(H)$ contains a directed
cycle consisting of exactly one read-dependency edge by $x$ from $T_j$
to $T_i$ and a set of edges by $y$ containing at least one
anti-dependency edge from $T_i$ to $T_j$ and $T_j$'s read from $y$
precedes its read from $x$.
\end{definition}

\begin{figure}[H]
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)\\
T_2 &: w_x(2)~w_y(2)\\
T_3 &: r_x(2)~r_y(1)
\end{align*}
\caption{Example of \textit{N-TA} anomaly.}
\label{fig:nta-history}
\end{figure}

@@

\begin{definition}[Read Atomic (RA)]
A system that provides Read Atomic isolation prohibits phenomenon
N-RA in addition to providing Read Committed isolation.
\end{definition}

@@

\begin{definition}[Writes Do Not Follow Reads (N-WFR)]
A history $H$ exhibits phenomenon N-WFR if, in $DSG(H)$, for all
committed transactions $T_1$, $T_2$, $T_3$ such that $T_2$
write-depends on $T_1$ and $T_3$ write-depends on $T_2$, $T_3$ does
not directly anti-depend on $T_1$.
\end{definition}

\begin{figure}[H]
\begin{align*}
\small
T_1 &: w_x(1)\\
T_2 &: r_x(1) w_y(1)\\
T_3 &: r_y(1)~r_x(0)
\end{align*}
\caption{Example of \textit{N-WFR} anomaly.}
\label{fig:nwfr-history}
\end{figure}

@@

\begin{definition}[Writes Follow Reads (WFR)]
A system provides Writes Follow Reads if it prohibits phenomenon N-WFR.
\end{definition}

\begin{definition}[Missing Your Writes (N-RYW)]
A history $H$ exhibits phenomenon N-RYW if $DSG(H)$ contains a directed cycle
consisting of a transitive session-dependency between transactions
$T_j$ and $T_i$, at least one anti-dependency edge, and the remainder
anti-dependency or write-dependency edges.
\end{definition}

\begin{figure}[H]
\begin{align*}
\small
T_1 &: w_x(1)\\
T_2 &: r_x(0)\\
\end{align*}
\caption{Example of \textit{N-RYW} anomaly if $T_2$ directly session-depends on $T_1$.}
\label{fig:nryw-history}
\end{figure}

@@

\begin{definition}[Read Your Writes (RYW)]
A system provides Read Your Writes if it prohibits phenomenon N-RYW.
\end{definition}

\begin{definition}[PRAM Consistency]
A system provides PRAM Consistency if it prohibits phenomenon N-MR,
N-MW, and N-RYW.
\end{definition}

\begin{definition}[Causal Consistency]
A system provides Causal Consistency if it provides PRAM Consistency
and prohibits phenomenon N-WFR.
\end{definition}

@@

\begin{definition}[Snapshot Isolation]
A system that provides Snapshot Isolation prevents phenomena G0, G1a,
G1b, G1c, N-P-CI, N-RA, and Lost Update.
\end{definition}

@@

\section{Transactional Atomicity}

In this section, we provide additional information on our two-phase
Transactional Atomicity algorithm.

@@

\begin{algorithmic}
\caption*{\textbf{Algorithm} HAT Transactional Atomicity}\vspace{.5em}
\State{\textit{\textbf{Shared Data Types}}\\\vspace{.5em}
\texttt{timestamp}: unique, per transaction identifier\\
$\texttt{write}:[\texttt{key:}k, \texttt{value}:v, \texttt{timestamp}:ts, \texttt{set<key>}:sibs]$\\\hrulefill}\vspace{.5em}

>>>>

To reason about isolation anomalies, we use Adya's concept of a
conflict graph, which is composed of dependencies between
transaction. The definitions in this section are directly from Adya,
with two differences. First, we expand Adya's formalism to deal
with \textit{per-item} dependencies. Second, we define session
dependencies (Definition~\ref{def:sd1}).

@@

\begin{definition}[Directly Read-Depends{~\cite[Definition 2]{adya}}]
$T_j$ directly read-depends on transaction $T_i$ if it directly
item-read-depends or directly predicate-read-depends on $T_i$.
\end{definition}

@@

\begin{definition}[Directly Anti-Depends{~\cite[Definition 4]{adya}}]
Transaction $T_j$ directly anti-depends on transaction $T_i$ if it
directly item-anti-depends or directly predicate-anti-depends on
$T_i$.
\end{definition}

@@

\begin{definition}[Directly Write-Depends{~\cite[Definition 5]{adya}}]
A transaction $T_j$ directly write-depends on transaction $T_i$ if
$T_i$ directly write-depends by $x$ on $T_j$ for some item $x$.
\end{definition}

@@

The dependencies for a history $H$ form a graph called its Directed
Serialization Graph ($DSG(H)$). If $T_j$ directly write-depends on
$T_i$ by $x$, we draw $T_i\overset{ww_x}\longrightarrow T_j$. If $T_j$
read-depends on $T_i$ by $x$, we draw
$T_i \overset{rw_x}\longrightarrow T_j$. If $T_j$ directly
anti-depends on transaction $T_j$ by $x$, we draw
$T_i \overset{wr_x}\DashedArrow T_j$. If $T_j$ session-depends on
$T_i$ in session $S$, we draw $T_i \overset{S}\longrightarrow
T_j$~\cite[Definition 8]{adya}.

We also consider the Unfolded Serialization Graph ($USG(H)$) that is a
variation of the $DSG$. The USG is speciﬁed for the transaction of
interest, $T_i$, and a history, $H$, and is denoted by $USG(H,
T_i)$. For the USG, we retain all nodes and edges of the $DSG$ except
for $T_i$ and the edges incident on it. Instead, we split the node for
$T_i$ into multiple nodes---one node for every read/write event in
$T_i$. The edges are now incident on the relevant event of $T_i$.

$USG(H, T_i)$ is obtained by transforming $DSG(H)$ as follows:.  For
each node $p$ ($p \neq T_i$) in $DSG(H)$, we add a node to
$USG(H,T_i)$. For each edge from node $p$ to node $q$ in $DSG(H)$,
where p and q are different from $T_i$, we draw a corresponding edge
in $USG(H,T_i)$. Now we add a node corresponding to every read and
write performed by $T_i$. Any edge that was incident on $T_i$ in the
$DSG$ is now incident on the relevant event of $T_i$ in the
$USG$. Finally, consecutive events in $T_i$ are connected
by \textit{order edges}, e.g., if an action (e.g., SQL statement)
reads object $y_j$ and immediately follows a write on object $x$ in
transaction $T_i$, we add an order-edge from $w_i(x_i)$ to
$r_i(y_j)$~\cite[Section 4.2.1]{adya}.

@@

\begin{definition}[Item-Many-Preceders (IMP)]
\label{def:imp}
A history $H$ exhibits phenomenon IMP if $DSG(H)$ contains a
transaction $T_i$ such that $T_i$ directly item-read-depends by $x$ on more
than one other transaction.
\end{definition}

\begin{figure}[H]
\begin{align*}
\small
T_1 &: w_x(1)\\
T_2 &: w_x(2)\\
T_3 &: r_x(1)~r_x(2)
\end{align*}
\caption{Example of \textit{IMP} anomaly.}
\label{fig:nici-history}
\end{figure}

@@

\begin{definition}[Item Cut Isolation (I-CI)]
A system that provides Item Cut Isolation prohibits phenomenon IMP.
\end{definition}

\begin{definition}[Predicate-Many-Preceders (PMP)]
A history $H$ exhibits phenomenon PMP if, for all predicate-based
reads $r_i(P_i:Vset(P_i))$ and $r_j(P_j:Vset(P_j)$ in $T_k$ such that
the logical ranges of $P_i$ and $P_j$ overlap (call it $P_o$), the set
of transactions that change the matches of $P_o$ for $r_i$ and $r_j$
differ.
\end{definition}

\begin{definition}[Predicate Cut Isolation (P-CI)]
A system that provides Predicate Cut Isolation prohibits phenomenon PMP.
\end{definition}

\begin{definition}[Observed Transaction Vanishes (OTV)]
A history $H$ exhibits phenomenon OTV if $USG(H)$ contains a directed
cycle consisting of exactly one read-dependency edge by $x$ from $T_j$
to $T_i$ and a set of edges by $y$ containing at least one
anti-dependency edge from $T_i$ to $T_j$ and $T_j$'s read from $y$
precedes its read from $x$.
\end{definition}

\begin{figure}[H]
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)\\
T_2 &: w_x(2)~w_y(2)\\
T_3 &: r_x(2)~r_y(1)
\end{align*}
\caption{Example of \textit{OTV} anomaly.}
\label{fig:nta-history}
\end{figure}

@@

\begin{definition}[Read Atomic (RA)]
A system that provides Read Atomic isolation prohibits phenomenon
OTV in addition to providing Read Committed isolation.
\end{definition}

@@

\begin{definition}[Missing Read-Write Dependency (MRWD)]
A history $H$ exhibits phenomenon MRWD if, in $DSG(H)$, for all
committed transactions $T_1$, $T_2$, $T_3$ such that $T_2$
write-depends on $T_1$ and $T_3$ write-depends on $T_2$, $T_3$ does
not directly anti-depend on $T_1$.
\end{definition}

\begin{figure}[H]
\begin{align*}
\small
T_1 &: w_x(1)\\
T_2 &: r_x(1) w_y(1)\\
T_3 &: r_y(1)~r_x(0)
\end{align*}
\caption{Example of \textit{MRWD} anomaly.}
\label{fig:nwfr-history}
\end{figure}

@@

\begin{definition}[Writes Follow Reads (WFR)]
A system provides Writes Follow Reads if it prohibits phenomenon MWRD.
\end{definition}

\begin{definition}[Missing Your Writes (MYR)]
A history $H$ exhibits phenomenon MYR if $DSG(H)$ contains a directed cycle
consisting of a transitive session-dependency between transactions
$T_j$ and $T_i$, at least one anti-dependency edge, and the remainder
anti-dependency or write-dependency edges.
\end{definition}

\begin{figure}[H]
\begin{align*}
\small
T_1 &: w_x(1)\\
T_2 &: r_x(0)\\
\end{align*}
\caption{Example of \textit{MYR} anomaly if $T_2$ directly session-depends on $T_1$.}
\label{fig:nryw-history}
\end{figure}

@@

\begin{definition}[Read Your Writes (RYW)]
A system provides Read Your Writes if it prohibits phenomenon MYR.
\end{definition}

\begin{definition}[PRAM Consistency]
A system provides PRAM Consistency if it prohibits phenomenon N-MR,
N-MW, and MYR.
\end{definition}

\begin{definition}[Causal Consistency]
A system provides Causal Consistency if it provides PRAM Consistency
and prohibits phenomenon MWRD.
\end{definition}

@@

\begin{definition}[Snapshot Isolation]
A system that provides Snapshot Isolation prevents phenomena G0, G1a,
G1b, G1c, PMP, OTV, and Lost Update.
\end{definition}

@@

\section{Read Atomic Isolation}

In this section, we provide additional information on our two-phase
algorithm for providing Read Atomic isolation.

@@

\begin{algorithmic}
\caption*{\textbf{Algorithm} HAT Read Atomic Isolation}\vspace{.5em}
\State{\textit{\textbf{Shared Data Types}}\\\vspace{.5em}
\texttt{timestamp}: unique, per transaction identifier\\
$\texttt{write}:[\texttt{key:}k, \texttt{value}:v, \texttt{timestamp}:ts, \texttt{set<key>}:sibs]$\\\hrulefill}\vspace{.5em}

### 2013-07-31 writeup/hat-vldb.tex
Editing pass
Removing a few references, squeezing space
Last round of comments from Alan and Aaron

\documentclass{vldb}
\let\proof\relax
\let\endproof\relax
%\usepackage{draftwatermarkbottom, draftwatermarktop}
\usepackage{scrextend, MnSymbol}

@@

@@

\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz, amsthm, float}
\usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
\newcommand{\subparagraph}{}

@@

\titlespacing{\section}{1ex}{2ex}{1ex}

\titlespacing{\subsection}{1ex}{1ex}{0.25ex}
\titlespacing{\subsubsection}{1ex}{1ex}{0.25ex}

@@

\begin{abstract}
\noindent To minimize network latency and remain online during server
failures and network partitions, many modern distributed data storage
systems eschew transactional functionality, which provides strong
semantic guarantees for groups of multiple operations over multiple
data items. In this work, we consider the problem of providing Highly
Available Transactions (HATs): transactional guarantees that do not
suffer unavailability during system partitions or incur high network
latency.  We introduce a new taxonomy of highly available systems and
analyze existing ACID and distributed consistency guarantees to
identify which can and cannot be achieved in HAT systems. This unifies
the literature on weak isolation, replica consistency, and highly
available systems. We analytically and experimentally quantify the
availability and performance benefits of HATs---often two to three
orders of magnitude over wide-area networks---as well as their
necessary semantic compromises.
\end{abstract}\vspace{-.5em}

>>>>

\documentclass{vldb}
\let\proof\relax
\let\endproof\relax
%\usepackage{draftwatermarkbottom, draftwatermarktop}
\usepackage{scrextend, MnSymbol}
\usepackage[hyphens]{url}

@@

\usepackage{breakurl}

@@

\usepackage{times, graphicx, xcolor, multicol, minipage, subfig, tikz, amsthm, float}
\usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)
\newcommand{\subparagraph}{}

@@

\titlespacing{\section}{1ex}{1ex}{0.25ex}

\titlespacing{\subsection}{1ex}{0.5ex}{0.25ex}
\titlespacing{\subsubsection}{1ex}{0.25ex}{0.25ex}

@@

\begin{abstract}
\noindent To minimize network latency and remain online during server
failures and network partitions, many modern distributed data storage
systems eschew transactional functionality, which provides strong
semantic guarantees for groups of multiple operations over multiple
data items. In this work, we consider the problem of providing Highly
Available Transactions (HATs): transactional guarantees that do not
suffer unavailability during system partitions or incur high network
latency.  We introduce a new taxonomy of highly available systems and
analyze existing ACID and distributed consistency guarantees to
identify which can and cannot be achieved in HAT systems. This unifies
the literature on weak transactional isolation, replica consistency,
and highly available systems. We analytically and experimentally
quantify the availability and performance benefits of HATs---often two
to three orders of magnitude over wide-area networks---as well as
their necessary semantic compromises.
\end{abstract}

### 2013-07-31 writeup/intro.tex
Last round of comments from Alan and Aaron

The last decade has seen a shift in the design of popular large-scale
database systems, from the use of transactional
RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
widespread adoption of loosely consistent distributed key-value
stores~\cite{bigtable, pnuts, dynamo}. Core to this shift was the 2000
introduction of Brewer's CAP Theorem, which stated that a highly
available system cannot provide ``strong'' consistency guarantees in
the presence of network partitions~\cite{brewer-slides}. Yet, as
formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability: the ability to read the most recent write to a data
item that is replicated across servers~\cite{herlihy-art}. Yet the
implications of CAP---which concern a fairly narrow form of
distributed data consistency---have often been conflated with the
ability to provide ACID database
properties~\cite{hat-hotos,brewer-slides, foundation-article}; this
misunderstanding has led to substantial confusion regarding replica
consistency, transactional isolation, and high availability. The
recent resurgence of transactional systems like Spanner~\cite{spanner}
suggests that programmers value transactional semantics, but,
unfortunately, existing transactional data stores do not provide
availability in the presence of partitions~\cite{foundation-article,
  hstore,spanner,eiger,  walter, calvin}.

@@

Our main contributions in this paper are as follows. We fill the gap
between the many previously proposed isolation and consistency models
and the goal of high availability.  We classify which among the wide
array of models are achievable with high availability, denoting them
as {\em Highly Available Transactions} (HATs). In doing so, we
demonstrate that although many implementations of HAT semantics are
not highly available, this is an artifact of the implementations
rather than an inherent property of the semantics. Our investigation
shows that, besides serializability, Snapshot Isolation and Repeatable
Read isolation are not HAT-compliant, while most other isolation
levels are achievable with high availability. We also demonstrate that
many weak replica consistency models from distributed systems are both
HAT-compliant and simultaneously achievable with several ACID
properties.

Our investigation is based on both impossibility results and several
constructive, proof-of-concept algorithms. For example, Snapshot
Isolation and Repeatable Read isolation are not HAT-compliant because
they require detecting conflicts between concurrent updates (as needed
for preventing Lost Updates or Write Skew phenomena), which we show is
unavailable. However, Read Committed isolation, transactionally atomic
ioslation, and many of the weak consistency models from database and
distributed systems are achievable via algorithms that rely on
multi-versioning and limited client-side caching. For several
guarantees, such as causal consistency with phantom prevention and
ANSI Repeatable Read, we consider a modified form of high availability
in which clients ``stick to'' (i.e., have affinity with) at least one
server---a property which is often implicit in the distributed systems
literature~\cite{herlihy-art, eiger, cac} but which requires explicit
consideration in a client-server replicated database context. This
sticky availability is widely employed~\cite{eiger, vogels-defs} but
is less restrictive (and therefore more easily achievable) than
traditional high availability.

At a high level, the virtues of HATs are guaranteed responses from any
replica, low latency, and a range of semantic guarantees including
``snapshot reads'' to transactional atomicity. However, highly
available systems are fundamentally unable to prevent concurrent
updates to shared data items and cannot provide recency guarantees for
reads. To understand when these virtues and limitations are relevant
in practice, we survey both practitioner accounts and academic
literature, perform experimental analysis on modern cloud
infrastructure, and analyze representative applications for their
semantic requirements. Our experiences with a HAT prototype running
across multiple geo-replicated datacenters indicate that HATs offer a
one to three order of magnitude latency decrease compared to
traditional distributed serializability protocols, and they can
provide acceptable semantics for a wide range of programs, especially
those with monotonic logic and commutative updates~\cite{calm,
  crdt}. HAT systems can also enforce arbitrary foreign key
constraints for multi-item updates and, in some cases, provide limited
uniqueness guarantees. However, particularly for programs with
non-monotonic logic, HATs can fall short, requiring
sometimes-unavailable mechanisms that are more
coordination-intensive. Overall, this paper quantifies the benefits
that Highly Available Transactions offer as well as the necessary
restrictions that they place on applications.

>>>>

The last decade has seen a shift in the design of popular large-scale
database systems, from the use of transactional
RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
widespread adoption of loosely consistent distributed key-value
stores~\cite{bigtable, pnuts, dynamo}. Core to this shift was the 2000
introduction of Brewer's CAP Theorem, which stated that a highly
available system cannot provide ``strong'' consistency guarantees in
the presence of network partitions~\cite{brewer-slides}. As formally
proven~\cite{gilbert-cap}, the CAP Theorem pertains to a data
consistency model called linearizability, or the ability to read the
most recent write to a data item that is replicated across
servers~\cite{herlihy-art}. However, despite its narrow scope, the CAP
Theorem is often misconstrued as a broad result regarding the ability
to provide ACID database properties with high
availability~\cite{hat-hotos,brewer-slides, foundation-article}; this
misunderstanding has led to substantial confusion regarding replica
consistency, transactional isolation, and high availability. The
recent resurgence of transactional systems like Spanner~\cite{spanner}
suggests that programmers value transactional semantics, but,
unfortunately, existing transactional data stores do not provide
availability in the presence of partitions~\cite{orleans,foundation-article,
  hstore,spanner,eiger, walter,calvin}.

@@

Our main contributions in this paper are as follows. We relate the
many previously proposed isolation and consistency models to the goal
high availability, which guarantees a response from every data replica
in the presence of arbitrary partitions between them.  We classify
which among the wide array of models are achievable with high
availability, denoting them as {\em Highly Available Transactions}
(HATs). In doing so, we demonstrate that although many implementations
of HAT semantics are not highly available, this is an artifact of the
implementations rather than an inherent property of the semantics. Our
investigation shows that, besides serializability, Snapshot Isolation
and Repeatable Read isolation are not HAT-compliant, while most other
isolation levels are achievable with high availability. We also
demonstrate that many weak replica consistency models from distributed
systems are both HAT-compliant and simultaneously achievable with
several ACID properties.

Our investigation is based on both impossibility results and several
constructive, proof-of-concept algorithms. For example, Snapshot
Isolation and Repeatable Read isolation are not HAT-compliant because
they require detecting conflicts between concurrent updates (as needed
for preventing Lost Updates or Write Skew phenomena), which we show is
unavailable. However, Read Committed isolation, transactional
atomicity (Read Atomic isolation), and many of the weak consistency
models from database and distributed systems are achievable via
algorithms that rely on multi-versioning and limited client-side
caching. For several guarantees, such as causal consistency with
phantom prevention and ANSI Repeatable Read, we consider a modified
form of high availability in which clients ``stick to'' (i.e., have
affinity with) at least one server---a property which is often
implicit in the distributed systems literature~\cite{herlihy-art,
  eiger, cac} but which requires explicit consideration in a
client-server replicated database context. This sticky availability is
widely employed~\cite{eiger, vogels-defs} but is a less restrictive
model (and therefore more easily achievable) than traditional high
availability.

At a high level, the virtues of HATs are guaranteed responses from any
replica, low latency, and a range of semantic guarantees including
several whose usefulness is widely acepted such as Read
Committed. However, highly available systems are fundamentally unable
to prevent concurrent updates to shared data items and cannot provide
recency guarantees for reads. To understand when these virtues and
limitations are relevant in practice, we survey both practitioner
accounts and academic literature, perform experimental analysis on
modern cloud infrastructure, and analyze representative applications
for their semantic requirements. Our experiences with a HAT prototype
running across multiple geo-replicated datacenters indicate that HATs
offer a one to three order of magnitude latency decrease compared to
traditional distributed serializability protocols, and they can
provide acceptable semantics for a wide range of programs, especially
those with monotonic logic and commutative updates~\cite{calm,
  crdt}. HAT systems can also enforce arbitrary foreign key
constraints for multi-item updates and, in some cases, provide limited
uniqueness guarantees. However, particularly for programs with
non-monotonic logic, HATs can fall short, requiring
sometimes-unavailable mechanisms that are more
coordination-intensive. Overall, this paper quantifies the benefits
that Highly Available Transactions offer as well as the necessary
restrictions that they place on applications.

### 2013-07-31 writeup/model.tex
Cutting down length
Last round of comments from Alan and Aaron

Traditionally, a system provides {\textbf{high availability}} if every
user that can contact a correct (non-failing) server eventually
receives a response from that server, even in the presence of
arbitrary, indefinitely long network partitions between
servers~\cite{gilbert-cap}.\footnote{Under this definition from the
  distributed systems literuter, systems that require a majority of
  servers to be online is not available. Similarly, a system which
  ensures that clients receive a response with high (but not perfect)
  probability is not available. This is a stringent but matches the
  assumptions made in the CAP Theorem~\cite{gilbert-cap} and
  guarantees low latency~\cite{abadi-pacelc}.} As in a standard
distributed database, designated servers might perform operations for
different data items; a server that can handle an operation for a
given data item is called a \textit{replica} for that
item.\footnote{There is a further distinction between a \textit{fully
    replicated} system, in which all servers are replicas for all data
  items, and a \textit{partially replicated} system, in which at least
  one server acts as a replica for only a subset of all data
  items. For generality, and, for applicability to many modern
  ``sharded'' or ``partitioned'' data storage systems~\cite{ bigtable,
    pnuts, spanner, dynamo, hstore}, we consider partially replicated
  systems.}

\subsection{Sticky Availability}

In addition to high availability, which allows operations on any
replica, distributed algorithms often assume a model in which clients
always contact the same logical (sets of) replica(s) across subsequent
operations. As we will discuss in Section~\ref{sec:hats}, clients can
ensure continuity between operations (e.g., reading their prior
updates to a data item) by maintaining affinity or ``stickiness'' with
a server or set of servers~\cite{vogels-defs}. In a fully-replicated
system, where all servers are replicas for all data items, stickiness
is simple: a client can maintain stickiness by contacting the same
server for each of its requests. However, to stay ``sticky'' in a
partially-replicated system, where servers are replicas for subsets of
the set of data items (which we consider in this paper), a client must
maintain stickiness with a single \textit{logical} server that may
consist of multiple physical servers.. We say that a system provides
\textbf{sticky availability} if, whenever a client's operation on an
item is executed against a logical server that has observed all of its
prior operations, it eventually receives a response, even in the
presence of indefinitely long partitions. A client may choose to
become sticky available by acting as a server itself; for example, a
client might cache its reads and writes~\cite{bolton,
  sessionguarantees, swift}. Any guarantee achievable in a highly
available system is achievable in a sticky high availability system
but not vice-versa.

>>>>

Traditionally, a system provides {\textbf{high availability}} if every
user that can contact a correct (non-failing) server eventually
receives a response from that server, even in the presence of
arbitrary, indefinitely long network partitions between
servers~\cite{gilbert-cap}.\footnote{Under this definition from the
  distributed systems literature, systems that require a majority of
  servers to be online are not available. Similarly, a system which
  ensures that clients receive a response with high (but not perfect)
  probability is not available. This is a stringent requirement but it
  matches the assumptions made in the CAP Theorem~\cite{gilbert-cap}
  and guarantees low latency~\cite{abadi-pacelc}.} As in a standard
distributed database, designated servers might perform operations for
different data items; a server that can handle an operation for a
given data item is called a \textit{replica} for that
item.\footnote{There is a further distinction between a \textit{fully
    replicated} system, in which all servers are replicas for all data
  items, and a \textit{partially replicated} system, in which at least
  one server acts as a replica for only a subset of all data
  items. For generality, and, for applicability to many modern
  ``sharded'' or ``partitioned'' data storage systems~\cite{ bigtable,
    pnuts, spanner, dynamo, hstore}, we consider partially replicated
  systems.}

\subsection{Sticky Availability}
\label{sec:sticky}

In addition to high availability, which allows operations on any
replica, distributed algorithms often assume a model in which clients
always contact the same logical (sets of) replica(s) across subsequent
operations. As we will discuss in Section~\ref{sec:hats}, clients can
ensure continuity between operations (e.g., reading their prior
updates to a data item) by maintaining affinity or ``stickiness'' with
a server or set of servers~\cite{vogels-defs}. In a fully-replicated
system, where all servers are replicas for all data items, stickiness
is simple: a client can maintain stickiness by contacting the same
server for each of its requests. However, to stay ``sticky'' in a
partially-replicated system, where servers are replicas for subsets of
the set of data items (which we consider in this paper), a client must
maintain stickiness with a single \textit{logical} server that may
consist of multiple physical servers. We say that a system provides
\textbf{sticky availability} if, whenever a client's operation on an
item is executed against a logical server that has observed all of its
prior operations, it eventually receives a response, even in the
presence of indefinitely long partitions. A client may choose to
become sticky available by acting as a server itself; for example, a
client might cache its reads and writes~\cite{bolton,
  sessionguarantees, swift}. Any guarantee achievable in a highly
available system is achievable in a sticky high availability system
but not vice-versa.
